---
title: "zmisc"
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file. -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)

# Package metadata, read with base R so that knitting needs no extra packages
pkg <- read.dcf("DESCRIPTION")[1, ]
squish <- function(x) gsub("\\s+", " ", trimws(x))
```

<!-- badges: start -->
[![CRAN version](https://www.r-pkg.org/badges/version/zmisc)](https://CRAN.R-project.org/package=zmisc)
[![GitHub version](https://img.shields.io/badge/Git-`r pkg[["Version"]]`-success)](https://github.com/torfason/zmisc)
[![R-CMD-check](https://github.com/torfason/zmisc/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/torfason/zmisc/actions/workflows/R-CMD-check.yaml)
<!-- badges: end -->

## `r squish(pkg[["Title"]])`

`r squish(pkg[["Description"]])`

## Installation

You can install the released version of `zmisc` from
[CRAN](https://cran.r-project.org/package=zmisc) with:

```r
install.packages("zmisc")
```

You can use `pak` to install the development version of `zmisc` from
[GitHub](https://github.com/torfason/zmisc) with:

```r
pak::pak("torfason/zmisc")
```

## Usage

In order to use the package, you generally want to attach it first:

```{r load_package}
library(zmisc)
```


## Quick and easy value lookups

The functions [lookup()] and [lookuper()] are used to look up values from a
lookup table, which can be supplied as a `vector`, a `list`, or a `data.frame`.
The functions are in some ways similar to the Excel function `VLOOKUP()`, but
are designed to work smoothly in an R workflow, in particular within pipes.

### lookup: Lookup values from a lookup table

The [lookup()] function implements lookup of values (such as variable names)
from a lookup table which maps keys onto values (such as variable labels or
descriptions).

The lookup table can be in the form of a two-column `data.frame`, in the form of
a named `vector`, or in the form of a `list`. If the table is in the form of a
`data.frame`, the key column should be named either `key` or `name`, and the
value column should be named `value` (for the value). If the lookup table is in
the form of a named `vector` or `list`, the names are used as the key, and the
returned value is taken from the values in the vector or list.

The underlying lookup is done using `base::match()`, and all atomic data types
except `factor` are supported. Factors are omitted due to the ambiguity in what
should be looked up (the values or the levels). It is important that `x`,
`.default` and the columns of `lookup_table` are all of the same type
(specifically of the same `base::mode()`). If the lookup table is specified as a
`vector` or `list`, only the `character` variables are supported, because
`name(lookup_table)` is always of mode `character`.

Original values are returned if they are not found in the lookup table.
Alternatively, a `.default` can be specified for values that are not found. Note
that it is possible to specify `NA` as one of the keys to look up NA values
(only when using a `data.frame` as lookup table).

Any names or attributes of x are preserved.

#### Examples

```{r lookup}
fruit_lookup_vector <- c(a = "Apple", b = "Banana", c = "Cherry")
lookup(letters[1:5], fruit_lookup_vector)
lookup(letters[1:5], fruit_lookup_vector, .default = NA)

mtcars_lookup_data_frame <- data.frame(
  name = c("mpg", "hp", "wt"),
  value = c("Miles/(US) gallon", "Gross horsepower", "Weight (1000 lbs)"))
lookup(names(mtcars), mtcars_lookup_data_frame)

# A more complex example, with numeric and NA values
numeric_lookup_table <- data.frame(
  key = c(1:5, NA), value = c(sqrt(1:5), 99999))
lookup(c(0:6, NA), numeric_lookup_table)
```

### lookuper: Construct lookup function based on a specific lookup table

The [lookuper()] function returns *a function* equivalent to the [lookup()]
function, except that instead of taking a lookup table as an argument, the
lookup table is embedded in the function itself.

This can be very useful, in particular when using the lookup function as an
argument to other functions that expect a function which maps
`character`->`character` (or other data types), but do not offer a good way to
pass additional arguments to that function.

#### Examples

```{r lookuper}
lookup_fruits <- lookuper(list(a = "Apple", b = "Banana", c = "Cherry"))
lookup_fruits(letters[1:5])

lookup_fruits_nomatch_na <-
  lookuper(list(a = "Apple", b = "Banana", c = "Cherry"), .default = NA)
lookup_fruits_nomatch_na(letters[1:5])
```


## Safer sampling, sequencing and aggregation

The functions [zample()], [zeq()], and [zingle()] are intended to make your code
less likely to break in mysterious ways when you encounter unexpected boundary
conditions. The [zample()] and [zeq()] are almost identical to the [sample()]
and [seq()] functions, but a bit safer.

### zample: Sample from a vector in a safe way

The [zample()] function duplicates the functionality of [sample()], with the
exception that it does not attempt the (sometimes dangerous) user-friendliness
of switching the interpretation of the first element to a number if the length
of the vector is 1. `zample()` *always* treats its first argument as a vector
containing elements that should be sampled, so your code won't break in
unexpected ways when the input vector happens to be of length 1.

#### Examples

```{r zample}
# For vectors of length 2 or more, zample() and sample() are identical
set.seed(42); zample(7:11)
set.seed(42); sample(7:11)

# For vectors of length 1, zample() will still sample from the vector,
# whereas sample() will "magically" switch to interpreting the input
# as a number n, and sampling from the vector 1:n.
set.seed(42); zample(7)
set.seed(42); sample(7)

# The other arguments work in the same way as for sample()
set.seed(42); zample(7:11, size = 13, replace = TRUE, prob = (5:1)^3)

# Of course, sampling more than the available elements without
# setting replace=TRUE will result in an error
set.seed(42); tryCatch(zample(7, size = 2), error = wrap_error)
```

### zeq: Generate sequence in a safe way

The [zeq()] function creates an increasing integer sequence, but differs from
the standard one in that it will not silently generate a decreasing sequence
when the second argument is smaller than the first. If the second argument is
one smaller than the first it will generate an empty sequence, if the difference
is greater, the function will throw an error.

#### Examples

```{r zeq}
# For increasing sequences, zeq() and seq() are identical
zeq(11, 15)
zeq(11, 11)

# If second argument equals first-1, an empty sequence is returned
zeq(11, 10)

# If second argument is less than first-1, the function throws an error
tryCatch(zeq(11, 9), error = wrap_error)
```

### zingle: Return the single (unique) value found in a vector

The [zingle()] function returns the first element in a vector, but only if all
the other elements are identical to the first one (the vector only has a
`zingle` value). If the elements are not all identical, it throws an error. The
vector must contain at least one non-`NA` value, or the function errors out as
well. This is especially useful in aggregations, when all values in a given
group should be identical, but you want to make sure.

#### Examples

```{r zingle}
# If all elements are identical, all is good.
# The value of the element is returned.
zingle(c("Alpha", "Alpha", "Alpha"))

# If any elements differ, an error is thrown
tryCatch(zingle(c("Alpha", "Beta", "Alpha")), error = wrap_error)
```

```{r zingle_dplyr}
if (require("dplyr", quietly = TRUE, warn.conflicts = FALSE)) {
  d <- data.frame(
    id    = c(1, 2, 1),
    name  = c("James", "Jack", "James"),
    fouls = c(3, 2, 4)
  )

  # If the data is of the correct format, all is good
  d |>
    group_by(id) |>
    summarise(name = zingle(name), total_fouls = sum(fouls))
}
```

```{r zingle_dplyr_error}
if (require("dplyr", quietly = TRUE, warn.conflicts = FALSE)) {
  # If a name does not match its ID, we should get an error
  d[1, "name"] <- "Jammes"
  tryCatch({
    d |>
      group_by(id) |>
      summarise(name = zingle(name), total_fouls = sum(fouls))
  }, error = wrap_error)
}
```


## Assertions with rlang-style errors

The `chk_*()` functions check the type and shape of an argument and, on failure,
raise an [rlang]-style error that names the argument as the caller wrote it. The
checking itself is done by [checkmate]. Each function returns its input, so an
assertion can sit in the middle of a pipe, and each is cheap enough on the
passing path to leave at the top of any function.

| R type       | Scalar            | Vector              |
| ------------ | ----------------- | ------------------- |
| `logical`    | `chk_flag()`      | `chk_logical()`     |
| `character`  | `chk_string()`    | `chk_character()`   |
| `numeric`    | `chk_number()`    | `chk_numeric()`     |
| `integer`    | `chk_inumber()`   | `chk_integer()`     |
| `double`     | `chk_dnumber()`   | `chk_double()`      |
| integerish   | `chk_znumber()`   | `chk_integerish()`  |
| naturalish   | `chk_count()`     | `chk_naturalish()`  |
| `factor`     |                   | `chk_factor()`      |
| `complex`    |                   | `chk_complex()`     |
| `raw`        |                   | `chk_raw()`         |
| `Date`       | `chk_day()`       | `chk_date()`        |
| `POSIXct`    | `chk_instant()`   | `chk_posixct()`     |
| any type     | `chk_scalar()`    | `chk_atomic()`      |

Every one of them takes the same small set of arguments rather than the argument
list of the [checkmate] function behind it: `na.ok`, `null.ok`, `attr.ok`, and
`length` or `range` where they apply, plus `zero.ok` for the naturalish pair.
The dots are reserved, so a misspelled argument raises rather than being quietly
ignored.

#### Examples

```{r chk_pipe}
# The assertion returns its input, so it composes in a pipe
c(2, 4, 6) |> chk_numeric(length = 3) |> sum()

# On failure, the error names the argument as the caller wrote it
my_mean <- function(x) {
  chk_numeric(x)
  sum(x) / length(x)
}
tryCatch(my_mean("seven"), error = wrap_error)
```

```{r chk_args}
# One parameter set across the whole family
tryCatch(chk_character(c("a", NA), na.ok = FALSE), error = wrap_error)
tryCatch(chk_integer(1:5, length = 3), error = wrap_error)
tryCatch(chk_numeric(c(1, 99), range = c(0, 10)), error = wrap_error)
tryCatch(chk_count(0, zero.ok = FALSE), error = wrap_error)

# attr.ok catches a vector that is not quite bare
labelled_ages <- structure(c(38L, 41L), label = "Age at interview")
tryCatch(chk_integer(labelled_ages), error = wrap_error)
```

Containers, classes and arbitrary conditions have their own smaller set, taking
`null.ok` and little else. `chk_true()` is the catch-all, for any property that
can be written as a condition.

```{r chk_other}
tryCatch(chk_list(data.frame(a = 1)), error = wrap_error)
tryCatch(chk_class(1:3, "factor"), error = wrap_error)
tryCatch(chk_true(nrow(mtcars) > 100), error = wrap_error)
```

Two [rlang] assertions are re-exported under the same naming convention.
`chk_dots_empty()` fails if anything was passed through `...`, and `chk_match()`
matches an argument against the values in its own default, returning the match.

```{r chk_match}
plot_kind <- function(kind = c("scatter", "line", "bar")) {
  kind <- chk_match(kind)
  kind
}
plot_kind()
tryCatch(plot_kind("pie"), error = wrap_error)
```


## Getting a better view on variables

The [notate()] function adds annotations to `factor` and `labelled` variables
that make it easier to see both values and labels/levels when using the [View()]
function.

### notate: Embed factor levels and value labels in values

This function adds level/label information as an annotation to either factors or
`labelled` variables. This function is called `notate()` rather than
`annotate()` to avoid conflict with `ggplot2::annotate()`. It is a generic that
can operate either on individual vectors or on a `data.frame`.

When printing `labelled` variables from a `tibble` in a console, both the
numeric value and the text label are shown, but no variable labels. When using
the `View()` function, only variable labels are shown but no value labels. For
factors, there is no way to view the integer levels and values at the same time.

In order to allow the viewing of both variable and value labels at the same
time, this function converts both `factor` and `labelled` variables to
`character`, including both numeric levels (`labelled` values) and character
values (`labelled` labels) in the output.

#### Examples

```{r notate}
d <- data.frame(
  chr = letters[1:4],
  fct = factor(c("alpha", "bravo", "chrly", "delta")),
  lbl = ll_labelled(c(1, 2, 3, NA),
                    labels = c(one = 1, two = 2),
                    label = "A labelled vector")
)
notate(d)
# View(notate(d))
```


[lookup()]:   https://torfason.github.io/zmisc/reference/lookup.html
[lookuper()]: https://torfason.github.io/zmisc/reference/lookuper.html

[zeq()]:      https://torfason.github.io/zmisc/reference/zeq.html
[zample()]:   https://torfason.github.io/zmisc/reference/zample.html
[zingle()]:   https://torfason.github.io/zmisc/reference/zingle.html

[notate()]:   https://torfason.github.io/zmisc/reference/notate.html

[checkmate]:  https://mllg.github.io/checkmate/
[rlang]:      https://rlang.r-lib.org/

[seq()]:      https://rdrr.io/r/base/seq.html
[sample()]:   https://rdrr.io/r/base/sample.html
[View()]:     https://rdrr.io/r/utils/View.html
