---
title: "zmisc"
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file. -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)

# Package metadata, read with base R so that knitting needs no extra packages
pkg <- read.dcf("DESCRIPTION")[1, ]
squish <- function(x) gsub("\\s+", " ", trimws(x))
```

<!-- badges: start -->
[![CRAN version](https://www.r-pkg.org/badges/version/zmisc)](https://CRAN.R-project.org/package=zmisc)
[![GitHub version](https://img.shields.io/badge/Git-`r pkg[["Version"]]`-success)](https://github.com/torfason/zmisc)
[![R-CMD-check](https://github.com/torfason/zmisc/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/torfason/zmisc/actions/workflows/R-CMD-check.yaml)
<!-- badges: end -->

## `r squish(pkg[["Title"]])`

`r squish(pkg[["Description"]])`

## Installation

Install the released version from
[CRAN](https://cran.r-project.org/package=zmisc):

```r
install.packages("zmisc")
```

Or the development version from
[GitHub](https://github.com/torfason/zmisc), using `pak`:

```r
pak::pak("torfason/zmisc")
```

## Usage

```{r load_package}
library(zmisc)
```


## Quick and easy value lookups

[lookup()] looks up values (such as variable names) in a lookup table that maps
keys onto values (such as variable labels). The table can be a two-column
`data.frame`, a named `vector`, or a `list`. Values that are not found are
returned unchanged, or replaced by `.default`.

```{r lookup}
fruit_lookup_vector <- c(a = "Apple", b = "Banana", c = "Cherry")
lookup(letters[1:5], fruit_lookup_vector)
lookup(letters[1:5], fruit_lookup_vector, .default = NA)

mtcars_lookup_data_frame <- data.frame(
  name = c("mpg", "hp", "wt"),
  value = c("Miles/(US) gallon", "Gross horsepower", "Weight (1000 lbs)"))
lookup(names(mtcars), mtcars_lookup_data_frame)
```

[lookuper()] returns *a function* that does the same lookup, with the table
embedded in the function itself. That is useful wherever a
`character`->`character` function is expected but there is no good way to pass
the table along with it.

```{r lookuper}
lookup_fruits <- lookuper(list(a = "Apple", b = "Banana", c = "Cherry"))
lookup_fruits(letters[1:5])
```


## Safer sampling, sequencing and aggregation

[zample()], [zeq()] and [zingle()] make code less likely to break in mysterious
ways at awkward boundary conditions.

[zample()] is [sample()] without the user-friendliness of switching to
`1:n` when the input happens to have length 1.

```{r zample}
# For vectors of length 2 or more, zample() and sample() are identical
set.seed(42); zample(7:11)
set.seed(42); sample(7:11)

# For length 1, sample() "magically" switches to sampling from 1:n
set.seed(42); zample(7)
set.seed(42); sample(7)
```

[zeq()] is an increasing integer sequence that refuses to silently run
backwards. A second argument one below the first gives an empty sequence,
anything lower is an error.

```{r zeq}
zeq(11, 15)
zeq(11, 10)
tryCatch(zeq(11, 9), error = wrap_error)
```

[zingle()] returns the single distinct value in a vector, and errors if there
is more than one. This is useful in aggregations where all values in a group
should be identical, but where the assumption is worth checking.

```{r zingle}
zingle(c("Alpha", "Alpha", "Alpha"))
tryCatch(zingle(c("Alpha", "Beta", "Alpha")), error = wrap_error)
```

```{r zingle_dplyr}
if (require("dplyr", quietly = TRUE, warn.conflicts = FALSE)) {
  d <- data.frame(
    id    = c(1, 2, 1),
    name  = c("James", "Jack", "James"),
    fouls = c(3, 2, 4)
  )

  d |>
    group_by(id) |>
    summarise(name = zingle(name), total_fouls = sum(fouls))
}
```


## Checks with rlang-style errors

The `chk_*()` functions check the type and shape of an argument and, on
failure, raise an [rlang]-style error naming the argument as the caller wrote
it. The checking itself is done by [checkmate]. Each function returns its
input, so a check can sit in the middle of a pipe, and each is cheap enough on
the passing path to leave at the top of any function.

```{r chk}
# The check returns its input, so it composes in a pipe
c(2, 4, 6) |> chk_numeric(length = 3) |> sum()

# On failure, the error names the argument as the caller wrote it
my_mean <- function(x) {
  chk_numeric(x)
  sum(x) / length(x)
}
tryCatch(my_mean("seven"), error = wrap_error)
```

There are checks for every atomic type in scalar and vector form, for lists and
composite objects, and for classes and arbitrary conditions. See
`vignette("chk")` for the full set and the shared parameters.


## Getting a better view on variables

[notate()] annotates `factor` and `labelled` variables so that both values and
labels are visible at once, which neither printing nor [View()] manages on its
own.

```{r notate}
d <- data.frame(
  chr = letters[1:4],
  fct = factor(c("alpha", "bravo", "chrly", "delta")),
  lbl = ll_labelled(c(1, 2, 3, NA),
                    labels = c(one = 1, two = 2),
                    label = "A labelled vector")
)
notate(d)
# View(notate(d))
```


## Other utilities

- [glue_vector()] applies a [glue()] template element-wise to a vector, for
  pipe-friendly interpolation outside a `data.frame`.
- [asciify()] transliterates accented characters to ASCII.
- [yencode()] encodes arbitrary strings into a restricted character set, and
  [ydecode()] reverses it.


[lookup()]:      https://torfason.github.io/zmisc/reference/lookup.html
[lookuper()]:    https://torfason.github.io/zmisc/reference/lookuper.html

[zeq()]:         https://torfason.github.io/zmisc/reference/zeq.html
[zample()]:      https://torfason.github.io/zmisc/reference/zample.html
[zingle()]:      https://torfason.github.io/zmisc/reference/zingle.html

[notate()]:      https://torfason.github.io/zmisc/reference/notate.html

[glue_vector()]: https://torfason.github.io/zmisc/reference/glue_vector.html
[asciify()]:     https://torfason.github.io/zmisc/reference/asciify.html
[yencode()]:     https://torfason.github.io/zmisc/reference/yencode.html
[ydecode()]:     https://torfason.github.io/zmisc/reference/yencode.html

[checkmate]:     https://mllg.github.io/checkmate/
[rlang]:         https://rlang.r-lib.org/
[glue()]:        https://glue.tidyverse.org/

[seq()]:         https://rdrr.io/r/base/seq.html
[sample()]:      https://rdrr.io/r/base/sample.html
[View()]:        https://rdrr.io/r/utils/View.html
